Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-06 paper/main.tex
more shortening for popl

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource consumption is closely related to the
number $A(1)$ of times the test \lstinline{a[i] > m} is true during the
execution.
%
We observe that $A(1)+1$ is the count of ``maximum candidates'' in the
array \lstinline{a} seen by the algorithm. We define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$.  $A(1)$ is
bounded by $N$, the size of the array.  So any automated tool
available would derive the linear bound $2 \cdot N$ for that program.
But with the expressivity of our logic it is possible to use the
previous set cardinal directly and precisely tie the bound to the
initial contents of the array.

The non-trivial part of this derivation is finding the loop
invariant for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximum candidates so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  % As a side remark, if $K \ge 0$, it is possible to
% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
% of the same program.  It can be done either by adapting the proof or by
% applying the {\sc L:Frame} rule one top of the triple already derived.
% More generally, without function calls, the {\sc L:Frame}
% rule is admissible in our system.

The example in~\pref{fig:xmplbs} shows a use case for logical
variables as well as a metric for stack consumption. In a stack
metric, we account a constant cost for a function call ($M_f>0$) that
is returned after the call ($M_r = -M_f <0$).  All other resource cost
are $0$.  We are interested in showing that a binary search function
$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
  + Z \cdot F}{}{Z \cdot F}$ to express that the stack required by the
function is fully returned afterwards.

The critical step in this proof is the application of the
{\sc L:Call} rule to prove the recursive call.  At this point
the context $\Delta$ contains the following specification
for the function \code{bsearch}
$$
  \forall y \, (x,l,h) \, \_.
  ( (\lambda y \, (\_, l, h) .\, (y {=} \log_2(h {-} l)) + y {\cdot} F)
  , (\lambda y \, \_ .\, y {\cdot} F)
  ).
$$
Using the {\sc L:Call} rule it is possible to instantiate
$y$ with $Z - 1$, and because $M_f = F$ and $M_r = -F$ in
the stack resource metric, the appropriate triple can be
derived on the call site.  The rest of the proof does not
involve any resource manipulation and is just bookkeeping
of logical assertions.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ (Z = \log_2(h-l)) + Z \cdot F \} }$
bsearch(x,l,h) {
  if (h-l > 1) {
    ${\color{blue}%
      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
    m = h + (h-l)/2;
    ${\color{blue}%
      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot F \} }$
    if (a[m]>x) h=m; else l=m;
    ${\color{blue}%
      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot F + F \} }$
    l = bsearch(x,l,h); }
  ${\color{blue}%
    \{ (Z-1) \cdot F - (-F) \} }$
  return l;
} ${\color{blue} \{ Z \cdot F \} }$
\end{lstlisting}
\caption{Example derivation of a stack usage bound for a binary
  search program.  The metric used sets a cost of $F$ before
  to function call and $-F$ after the call, $F$ is the stack
  frame size of the function $\code{bsearch}$.  We use $Z$ as
  logical variable.
  }
\label{fig:xmplbs}
\end{figure}

@@

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as
\begin{align*}
  \forall \state \, \cost \, m \, \cost' .\,&
  (m \le n \land P(\state) \le \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
  &\implies \cost' \ge 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used in the
soundness proof for the function-call and loop cases.  The constraint
$\cost' \geq 0$ imposed by the definition ensures that the program does
not stop because of a resource error (recall that a negative resource
counter on the right-hand side is a resource failure).  However, it
does not rule out memory safety errors or assertion failures. This is
because our logic does not prove any safety or correctness theorems
but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition of validity.

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we can define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.  A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

If $S$ is a complete program and $\Delta$ is empty, by
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$, so
we get
$$
\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resources for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.

@@

In this section we present the automatic amortized analysis that we
use in our implementation to derive linear resource bounds for C
programs.

@@

\sectskip
\bibliographystyle{abbrvnat}
\bibliography{lit}

>>>>

In the example in~\pref{fig:xmplmax} we derive a precise runtime
bound on a program that searches a maximal element in an array.  The
cost metric that we use simply counts the assignments performed by the
program.  Hence, the resource cost is closely related to the
number of times the test \lstinline{a[i] > m} is true during the
execution.
%
If we define
$$
 A(i) =  \# \{ k \mid i \le k < N \land \forall\, 0 \le j < k.\, a[j] < a[k] \} \, .
$$
where we write $\# S$ for the cardinal of the set $S$ then $A(1)+1$ is
the number of ``maximum candidates'' in the array \lstinline{a} seen
by the algorithm. $A(1)$ is bounded by $N$, the size of the array.
So any automated tool would at best derive the linear bound $2 \cdot
N$ for that program.  But with the expressivity of our logic it is
possible to use the previous set cardinal directly and precisely tie
the bound to the initial contents of the array.
%
The non-trivial part of this derivation is finding the loop invariant
$(m = \max_{k \in [0, i-1]} a[k]) + A(i) + (N-i)$ for the while loop.
When the condition \lstinline{a[i] > m} is true, we know that we
encountered a ``maximum so far'' because $m$ is a maximal element of
\lstinline{a[$0 \dots i$]}, thus $A(i) = 1 + A(i+1)$ and we get one
potential unit to pay for the assignment.  In the other case, no
maximum so far is encountered so $A(i) = A(i+1)$.
% As a side remark, if $K \ge 0$, it is possible to
% have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
% of the same program.  It can be done either by adapting the proof or by
% applying the {\sc L:Frame} rule one top of the triple already derived.
% More generally, without function calls, the {\sc L:Frame}
% rule is admissible in our system.

The example in~\pref{fig:xmplbs} shows a use case for logical
variables as well as a metric for stack consumption. In a stack
metric, we account a constant cost for a function call ($M_f>0$) that
is returned after the call ($M_r = -M_f <0$).  All other resource cost
are $0$.  We are interested in showing that a binary search function
$\code{bsearch}$ has logarithmic stack consumption.  We use a logical
variable $Z$ in the function specification $\htriple{(Z = \log_2(h-l))
  + Z \cdot M_\code{bsearch}}{}{Z \cdot M_\code{bsearch}}$ to express that the stack required by the
function is returned after the call.
%
The critical step in the proof is the application of the
{\sc L:Call} rule to the recursive call.  At this point
the context $\Delta$ contains the specification
$
  \forall y \, (x,l,h) \, \_.
  ( (\lambda y \, (\_, l, h) .\, (y = \log_2(h {-} l)) {+} y {\cdot} M_\code{bsearch})
  , (\lambda y \, \_ .\, y {\cdot} M_\code{bsearch})
  )
$.
%
Using the rule {\sc L:Call} it is possible to instantiate $y$ with $Z
- 1$, and because $M_f = M_\code{bsearch}$ and $M_r =
-M_\code{bsearch}$ in the stack metric.  The rest of the proof does
not involve any resource manipulation and is just bookkeeping of
logical assertions.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ (Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
bsearch(x,l,h) {
  if (h-l > 1) {
    ${\color{blue}%
      \{ (Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
    m = h + (h-l)/2;
    ${\color{blue}%
      \{ (m = \frac{h+l}{2} \land Z \ge 1 \land Z = \log_2(h-l)) + Z \cdot M_\code{bsearch} \} }$
    if (a[m]>x) h=m; else l=m;
    ${\color{blue}%
      \{ (Z - 1 = \log_2(h-l)) + (Z-1) \cdot M_\code{bsearch} + M_\code{bsearch} \} }$
    l = bsearch(x,l,h); }
  ${\color{blue}%
    \{ (Z-1) \cdot M_\code{bsearch} - (-M_\code{bsearch}) \} }$
  return l;
} ${\color{blue} \{ Z \cdot M_\code{bsearch} \} }$
\end{lstlisting}
\caption{Example derivation of a stack usage bound for a binary
  search program.  The used resource metric defines the cost $M_\code{bsearch}$ before
  the function call and $-M_\code{bsearch}$ after the call.  $M_\code{bsearch}$ is the stack
  frame size of the function $\code{bsearch}$.  $Z$ is a
  logical variable.
  }
\label{fig:xmplbs}
\end{figure}

@@

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as $\forall \state \, \cost \, m \, \cost' .$
\begin{align*}
  (m {\le} n \land P(\state) {\le} \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost'))
  \implies \cost' {\ge} 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used for
induction in the soundness proof for the function-call and loop cases.
The constraint $\cost' \geq 0$ imposed by the definition ensures that
the program does not stop because of a resource error (recall that a
negative resource counter on the right-hand side is a resource
failure).  However, it does not rule out memory safety errors or
assertion failures. This is because our logic does not prove any
safety or correctness theorems but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition\iffull{ of validity.}{.}

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  \iffull{We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}}
{
We define $\safeK n B R Q \cont \mathop{:=}
\safe n B {\code{break}} \cont \, \land \, 
\safe n Q {\code{skip}} \cont \, \land \, 
\safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont$.

}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$. \iffull{A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.}{}  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
We are now able to state the soundness of the quantitative logic.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable then
  $
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

Assume that $S$ is a complete program and $\Delta$ is empty.  By
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$. So
we derive
$$
\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resources for any run of the program $S$.  \iffull{This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.}{}

@@

In this section we present the automatic amortized analysis that we
use in our implementation to derive resource bounds for C
programs.

@@

\bibliographystyle{abbrvnat}
\iffull{
\bibliography{lit}
}
{
\sectskip
\footnotesize
\bibliography{lit}
}
